---
title: "RTI Data Science Exercise Steps 1 & 2"
output: html_notebook
---
<br>
<br>

##This Exercise is completed through 7 major steps: <br>
<br>

###**Step 1.** Read in and Stage data for Analysis <br>
###**Step 2.** High Level and Univariate Exploratory Analysis <br>
###**Step 4.** Split the data into Training, Validation, and Test sets <br>
###**Step 4.** Variable Associations With The Response <br>
###**Step 5.** Variable Transformation and Creation <br>
###**Step 6.** Model Building <br>
###**Step 7.** Model Comparison and Selection <br>


***
***

<br>
<br>
<br>

###Step 1. Read in and Stage data for Analysis

***
***

```{r, warning = F, message=F}
library(tidyverse)

# read in data
records <- read_csv('C:/Users/Will/Documents/MSA/jobs/rti_challenge/data-scientist-exercise01/data/flat_records.csv',
                    col_names = F)

# add column names
names(records) <- c('id', 'age', 'workclass_id', 'education_level_id', #records
                    'education_num', 'marital_status_id',  #records
                    'occupation_id','relationship_id', 'race_id', #records
                    'sex_id', 'capital_gain', 'capital_loss',   #records
                    'hours_week', 'country_id', 'over_50k', #records
                    'country_id2','country', #countries
                    'education_level_id2', 'education_level', #education_levels
                    'marital_status_id2', 'marital_status', #marital_statuses
                    'occupation_id2','occupation', #occupations
                    'race_id2', 'race', #races
                    'relationship_id2', 'relationship', #relationships
                    'sex_id2', 'sex', #sexes
                    'workclass_id2', 'workclass') #workclasses

# replace all ? with NA
records[records == '?'] <- NA

# Remove the id columns, convert character columns to factor,
# and re-order the education levels
records %>% 
  select(-contains('id')) %>%  
  mutate_each(funs(factor(.)), -c(age:hours_week)) %>% 
  mutate(education_level = factor(education_level, 
                                   levels =  c("Preschool", "1st-4th", "5th-6th", 
                                                "7th-8th", "9th", "10th", "11th",
                                                "12th", "HS-grad", "Some-college", 
                                                "Assoc-voc",  "Assoc-acdm", "Bachelors",
                                                "Masters","Prof-school", "Doctorate")))  -> records
```

```{r}
#check final dataset dimensions
dim(records)
```

***
***
<br>
<br>

###Step 2: High Level and Univariate Exploratory Analysis <br>
<br>

#### **Summary**

The Exploratory Analysis revealed some interesting insights about the data: <br>

1. The survey is most likely filled out by the men of the household: This is evident by the variables gender, and relationship.
2. There are a few variables that have some hard cutoff values including age, hours worked, and capital gain. 
3. There are people who report they have never worked, yet they report on average close to 25 hours a week.



<br>
<br>
<br>


####**High Level Data Overview**

I begin by looking at the first 5 rows, the last 5 rows, and a summary of each variable. I am just looking to understand the data at a high level here.

***

```{r}
head(records)
```

```{r}
tail(records)
```

```{r}
summary(records)
```


***
<br>
<br>

####**Univariate Distributions** <br>

Now I'm interested in learning more about each variable distribution. I want to identify any unepected patterns in the data and gain a better understanding of the data overall.

***

<br>
**age** <br>
This distribution looks as expected, except for the high count at 90. Potentially people are lying,
or the value was missing and for some reason was imputed as 90.

```{r}
  ggplot(records) + geom_histogram(aes(x = age), binwidth = 1)
```
<br>

**education_num** <br>
This distribution looks as expected
```{r}
ggplot(records) + geom_bar(aes(education_level)) + theme(axis.text.x = element_text(angle = 90, hjust = 1))
```
<br>

**capital_gain** 

```{r}
ggplot(records) + geom_histogram(aes(capital_gain))
```
<br>

**capital_gain without 0** <br>
It is odd that $99,999 has such a high frequency count. I'm also wondering why there isn't more variability here? I'm guessing most of the people in the united states aren't actively selling assets or property.. which makes this distribution more sensible. 
```{r}
records %>% 
  filter(capital_gain > 0)%>% 
  ggplot() + geom_histogram(aes(capital_gain))
```
<br>

**capital_loss**  <br>
This variable distribution looks as expected.
```{r}
  ggplot(records) + geom_histogram(aes(capital_loss))
```
<br>



**capital_loss without 0**
```{r}
records %>% 
  filter(capital_loss > 0)%>% 
  ggplot() + geom_histogram(aes(capital_loss))
```
<br>


**hours_week** <br>
It is a little odd that there is a hard cutoff here at 100 hours. The distribution makes sense overall though.
```{r}
  ggplot(records) + geom_bar(aes(x = hours_week))
```
<br>

**hours_week without 40**
```{r}
records %>% 
  filter(hours_week != 40) %>% 
  ggplot() + geom_bar(aes(x = hours_week))
```
<br>


**over_50k** <br>
This distribution looks as expected
```{r}
  ggplot(records) + geom_bar(aes(x = over_50k))
```
<br>



**country** <br>
This distribution looks as expected
```{r}
  ggplot(records) + geom_bar(aes(country)) + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```
<br>

**Marital Status** <br>
The 'never married' proportion is suprisingly high. This must be because because the census is at the household level. The number of 'married' people should be double the frequency, but since only one spouse responds to the survey that number is halved.
```{r}
ggplot(records) + geom_bar(aes(marital_status)) + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

**Occupation** <br>
Distribution looks as expected here. Maybe the NA categories consists mainly of unemployed people
```{r}
ggplot(records) + geom_bar(aes(occupation)) + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```
<br>




**race** <br> 
I would have thought the proportion of other races would be higher here.
```{r}
ggplot(records) + geom_bar(aes(race)) + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```
<br>


**relationship** <br>
Maybe the husband count is much higher than wife count because the males of each household are filling the census out.
```{r}
ggplot(records) + geom_bar(aes(relationship)) + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```
<br>

**sex** <br>
The gender difference here is not representative of the US population as a whole. Again, this can likely be occuring because the males of the houselholds are filling this out.
```{r}
ggplot(records) + geom_bar(aes(sex)) + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```
<br>


**workclass** <br>
  
This distribution looks as expected 
```{r}
ggplot(records) + geom_bar(aes(workclass)) + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

***
<br>
<br>

####**Bi-Variate Plots**
<br>
In this section I create visualizations in order to better understand associations between variables.

***

**Age vs. Work Hours** <br>
I am a little surprised that the number of working hours doesn't drop off more near age 75. 
```{r}
#here we impute age 90 with the median age because there was an abnormal amount of 
# 90's in this variable. It appeared as if 90 was the substitute for some missing, or
# that was the highest allowable age on the census and people who were recording that value
# were not being truthful

records %>% 
  mutate(age = ifelse(age == 90, median(age), age)) %>% #impute age 90 with median age
ggplot(aes(x=age, y=hours_week)) + geom_point(size = 8,alpha = .009) + geom_smooth()

```
 <br>
 <br>
 
**Education vs. Work Hours** <br>
Interesting here that the higher education one has, the more they are working. I guess the opportunity cost is higher if you aren't working with an advanced degree since the earning potential is much higher. It could also be that it is easier to get a job with a degree.
```{r} 
ggplot(records,aes(x=education_level, y=hours_week)) + geom_boxplot() + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

<br>
<br>
**Work Class vs. Work Hours** <br>
I thought federal and state government had strict 40 hour work weeks. It is weird that we are observing up to 100 hours in these sectors. It is odd that within the never worked category people are reporting working close to 25 hours a week on average.
```{r}
ggplot(records, aes(workclass,hours_week)) + geom_boxplot() + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

<br>
<br>
**Education Level vs. Marital Status** <br>
The pattern between marital status and education appears to be consistant
```{r}

table(records$marital_status, records$education_level) %>% 
  data.frame %>% 
  rename(marital_status = Var1, education_level = Var2) -> plot_df
  
ggplot(plot_df, aes(education_level,marital_status)) + geom_tile(aes(fill = sqrt(Freq)),colour = "white") + scale_fill_gradient(low = "white",high = "steelblue")+ 
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```


<br>
<br>
**Education Level vs. Capital Loss** 

```{r}
ggplot(records, aes(education_level,capital_loss)) + geom_boxplot() + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```
<br>
<br>

**Education Level vs. Capital Loss (Excluding 0)** <br>
Capital loss seems to be spread across education levels almost evenly
```{r}
records %>% 
  filter(capital_loss > 0) %>% 
ggplot(aes(education_level,capital_loss)) + geom_boxplot() + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

<br>
<br>

**Education vs. Capital Gain** 

```{r}
ggplot(records, aes(education_level,capital_gain)) + geom_boxplot() + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```
<br>
<br>

**Education vs. Capital Gain (Excluding 0)** <br>
As expected, the more education someone has, the more capital gains they are reporting
```{r}
records %>% 
  filter(capital_gain > 0) %>% 
ggplot(aes(education_level,capital_gain)) + geom_boxplot() + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```


***
***
